Decompositions of congruence subgroups \\ of Chevalley groups

### 2016-09-01 paper.tex
Fixed grant acknowledgement
Remastered

\documentclass[oneside, 12pt]{amsart}
\usepackage{amscd, amsmath, amssymb, amsthm, amsfonts, amstext, geometry, verbatim, enumitem, graphicx, mathtools, xfrac, tikz-cd, microtype, nameref, thmtools}
\usepackage[breaklinks=true]{hyperref}
\usepackage[capitalize]{cleveref}

@@

\oddsidemargin 5mm
\marginparwidth 5mm
\topmargin 0mm
\textheight 225mm
\textwidth 165mm
\headheight 0mm
\headsep 10mm
\footskip 5mm

@@

\title [Decompositions of congruence subgroups of Chevalley groups] {Decompositions of congruence subgroups \\ of Chevalley groups}
\keywords {parabolic decomposition, bounded generation, $\SL_2$-factorization, subsystem factorization, product decomposition. {\em Mathematical Subject Classification (2010):} 20G40, 20G35, 19B14}
\author{Sergey Sinchuk}
\author{Andrei Smolensky}
\email{sinchukss {\it at} yandex.ru; andrei.smolensky {\it at} gmail.com}
\thanks{Authors of the present paper acknowledge the financial support from Russian Science Foundation grant 14-11-00297}

@@

\begin{abstract} 
We formulate and prove relative versions of several classical decompositions known in the theory of Chevalley groups over commutative rings.
As an application we obtain upper estimates for the width of principal congruence subgroups in terms of several families of generators.
Some of our results are new even in the absolute case and were previously studied only for groups over finite fields.
\end{abstract}

@@

\section{Introduction}\label{sec:intro}
Let $\G(\Phi, R)$ be the simply-connected Chevalley group of type $\Phi$ over an arbitrary commutative ring $R$ and let $\E(\Phi, R)$ be its \emph{elementary subgroup},
i.\,e. the subgroup generated by the elementary root unipotents $x_\alpha(\xi)$, $\alpha\in\Phi$, $\xi\in R$, see~\cite{VP, St78, S}.
For an ideal $I \trianglelefteq R$ we denote by $\G(\Phi, R, I)$ the \emph{relative Chevalley group} (alias the \emph{principal congruence subgroup}) of level $I$, 
i.\,e. the kernel of the map $\G(\Phi, R)\rightarrow\G(\Phi, R/I)$ induced by the canonical projection $R\rightarrow R/I$.

We will also use the notion of the \emph{relative elementary Chevalley subgroup} $\E(\Phi, R, I) \leqslant \G(\Phi, R, I)$.
Recall that $\E(\Phi, R, I)$ is defined to be the normal closure in $\E(\Phi, R)$ of the subgroup generated by the set $\mathcal{X} = \{x_\alpha(\xi) \mid \alpha\in \Phi, \ \xi \in I \}$.

Set $z_\alpha(s, \xi) = x_{\alpha}(s)^{x_{-\alpha}(\xi)}$, $s\in I$, $\xi\in R$.
It is a classical result of J.~Tits and L.~Vaserstein (see~\cite{Tits76}, \cite[Theorem~2]{Va86}) that the relative elementary subgroup $\E(\Phi, R, I)$ is generated by
the set of elements $\mathcal{Z}(\Phi)$.

Here for a subset of roots $S\subseteq\Phi$ we denote by $\mathcal{Z}(S)$ the set
\[ \mathcal{Z}(S) =  \mathcal{X} \cup \{ z_\alpha(s, \xi) \mid s\in I, \ \xi \in R, \ \alpha \in S\}.\]

The following recent result of A.~Stepanov served as a starting point for the present paper.
The notation for the subgroups involved in the statements of the results below is introduced properly in section~\ref{sec:prelim}.
\begin{prop}[{\cite[Theorem~3.4]{S}}] \label{prop:Stepanov}
Let $\Phi$ be a reduced irreducible root system of rank $\geq 2$, let $S \subset \Phi$ be an arbitrary parabolic subset of roots, and $\Sigma_S$ be the special part of $S$. The relative elementary subgroup $\E(\Phi, R, I)$ is generated as an abstract group by the set $\mathcal{Z}(\Sigma_S)$. \end{prop}

The purpose of this paper is two-fold. 
First, we obtain relative analogues of several classical parabolic factorizations of Chevalley groups such as the Gauss, Bass---Kolster and Dennis---Vaserstein decompositions.
Then we apply these factorizations to the study of various bounded generation properties of Chevalley groups over rings.

Our first result is a relative version of the Gauss decomposition formulated in terms of the \emph{stable rank} of the ideal $I$ (see section~\ref{sec:stability-conditions} for the definition)
\begin{thm}\label{thm:srRI1}
Let $\Phi$ be a root system, let $I$ be an ideal of an arbitrary commutative ring $R$, and assume that $\sr(I)=1$.
Then the relative elementary Chevalley group $\E(\Phi, R, I)$ admits the following decomposition:
\[ \E(\Phi, R, I) = \Hh(\Phi, R, I) \cdot \U(\Phi, I) \cdot \U(\Phi^-, I) \cdot \U(\Phi, I). \]
\end{thm}
In the special case when $I$ is contained in the Jacobson radical of $R$, \cref{thm:srRI1} boils down to~\cite[Proposition~2.3]{Abe76}. 
We also note that the absolute case (i.\,e. $I=R$) of the theorem was treated by the second named author in~\cite{Sm12}.

The next main result is the relative Dennis---Vaserstein decomposition, which holds for a broader class of rings as compared to \cref{thm:srRI1}.
\begin{thm}\label{thm:DennisVaserstein}
Let $\Phi \neq \rE_8$ be a reduced irreducible root system of rank $\ell\geqslant 2$ and let $\{ r, s \}$ be a pair of distinct terminal vertices of the Dynkin diagram of $\Phi$. Denote the distance between $r$ and $s$ by $d$.
Assume that $\Phi$, $I$ and $\{r, s\}$ satisfy either of the following assumptions:
\begin{thmlist}
 \item $\sr(I) \leqslant d$ for classical $\Phi$,
 \item $\sr(I) \leqslant d$ for $\Phi=\rE_\ell$ with $\{r, s\} = \{ 2, \ell \}$,
 \item $\asr(I) \leqslant d$ for $\Phi=\rE_\ell$ with $\{r, s\} = \{1, \ell \}$,
 \item $\asr(I) \leqslant d$ for $\Phi=\rF_4$.
\end{thmlist}
Then $\E(\Phi, R, I) = \EP_r(R, I) \cdot \U(\Sigma^-_r \cap \Sigma^-_s, I) \cdot \EP_s(R, I).$ \end{thm}
Even in the special case $I=R$ the above theorem contains several new cases as compared to~\cite[Theorems~2.5 and~4.1]{St78}.
The term ``Dennis---Vaserstein decomposition'' is due to A.~Suslin and M.~Tulenbaev (cf.~\cite[Lemma~2.1]{ST76}).

The condition $\asr(I) \leq d$ used in the statement of \cref{thm:DennisVaserstein} is a relative version of the \emph{absolute stable range} condition of D.~Estes and J.~Ohm, which depends only on the ideal $I$ (see section~\ref{sec:rel-asr}).

Let $G$ be an abstract group with a generating set $X \subset G$. 
We denote by $W(G, X)$ the \emph{width of $G$ with respect to $X$}, i.\,e. the smallest natural number $N$ such that every element of $G$ is a product of at most $N$ elements of $X$.

Our first application of the parabolic factorizations is the following theorem, which gives an upper bound on the width of $\E(\Phi, R)$ in terms of Tits generators. 
Thus, it is a quantitative analogue of \cref{prop:Stepanov}
and a relative analogue of the results of~\cite{Tavgen91, VseUnitrZ1p, VavSmSuUnitrEng}.
\begin{thm}\label{thm:width} Let $I$ be an ideal of a commutative ring $R$.
\begin{thmlist}
\item If $R=\mathcal{O}_S$ is a Dedekind ring of arithmetic type posessing a real embedding and $\Phi$ is classical of rank $\geqslant2$, then 
$W(\G(\Phi, R, I), \mathcal{Z}(\Sigma_\ell^-))$ is finite;
\item If $\sr(I) = 1$ and $\Phi$ is an arbitrary irreducible root system, then 
\[W(\E(\Phi, R, I), \mathcal{Z}(\Pi))\leqslant 3|\Phi^+|+2\rk(\Phi)-1;\]
\item If $R = \mathbb{Z}[\sfrac{1}{p}]$ for some prime number $p$, then under the assumption of the Generalized Riemann Hypothesis one has
\begin{alignat*}{2}
& W(\E(\Phi, R, I), \mathcal{Z}(\Sigma_\ell^-))\leqslant 3|\Phi^+| + 2\rk(\Phi) + 1 & \text{ if } \Phi=\rA_\ell, \rC_\ell, \\
& W(\E(\Phi, R, I), \mathcal{Z}(\Sigma_\ell^-))\leqslant 4|\Phi^+| + \rk(\Phi) + 1 & \text{ if } \Phi=\rB_\ell, \rD_\ell.
\end{alignat*}
\end{thmlist}
\end{thm}
The first assertion of the above theorem is essentially a reformulation of a previously known result of O.~Tavgen (see \cite{TavgenThesis}, cf.~\cite{Mo07}), 
while for the second and the third assertion we give a direct proof.
Compare \cref{thm:width} with a recent result of U.~Hadad and D.~W.~Morris \cite[Theorem~1.6]{Ha12} which asserts that 
$W(\SL(n, \mathbb{Z}, I), \mathcal{Z}(\Pi))$ is finite for any ideal $I \leq \mathbb{Z}$ and $n \geq 3$.

Another application of the parabolic factorizations developed in the present paper is the subsystem factorization in terms of subgroups of type $\rA_1$.
The case of a finite field $R=F$ was studied by M.~Liebeck, N.~Nikolov and A.~Shalev in~\cite{LNS11}.
More precisely, for an untwisted Chevalley group it has been proved that $\G(\Phi, F)$ is a product of at most $N=5|\Phi^+|$ copies of $\SL(2, F)$.

In~\cite{V13} N.~Vavilov and E.~Kovach noted that the Bruhat decomposition immediately implies a bound of $N=3|\Phi^+|$ factors for an arbitrary field $F$.
In fact, in the assumptions of \cref{thm:srRI1} one can immediately obtain the decomposition of $\E(\Phi, R, I)$ with $3|\Phi^+|$ factors of type $\rA_1$.
Our next result further extends this to an even larger class of rings (which, for example, includes the ideals of all Dedekind rings).
\begin{thm}\label{thm:SL2width}
Let $I\trianglelefteq R$ be an ideal, $\Phi$ be an irreducible classical root system satisfying one of the following assumptions.
\[\begin{array}{l@{\quad}l@{\quad}l}
\Phi=\rA_\ell, & \sr(I)\leqslant 2, & N=3\left|\Phi^+\right| - \rk(\Phi) - 1; \\
\Phi=\rC_\ell, & \sr(I)\leqslant 3, & N=3\left|\Phi^+\right| - 2; \\
\Phi=\rB_\ell, \rD_\ell, & \asr(I)\leqslant 2, & N=4\left|\Phi^+\right| - 3\rk(\Phi).
\end{array}\]
Then the principal congruence subgroup $\G(\Phi, R, I)$ can be presented as a product of at most $N$ copies of its (regularly embedded) subgroups isomorphic to $\SL_2(R, I)$.
\end{thm}

Yet another application of the parabolic factorizations concerns the product decompositions of Chevalley groups formulated in terms of subgroups of type $\rA_\ell$ of submaximal rank.
Recall that in \cite{Nik07} N.~Nikolov proves the following result. 
\begin{thm*} Let $G$ be a classical Chevalley group (of normal or twisted type) of rank $\ell$ over a finite field $F$.
Then $G$ can be presented as a product of at most $200$ conjugates of its subgroups of type $\rA_{\ell-1}$. \end{thm*}

It is classically known that over an arbitrary field $F$ the group $\SL_{n+1}(F)$, $n\geq 3$ is a product of at most $4$ subgroups of type $\SL(n, F)$.
On the other hand, one immediately obtains from \cref{thm:DennisVaserstein} and surjective stability for linear $\K_1$-functor that $\SL_{n+1}(R, I)$ factors into a product
of a most $5$ subgroups of type $\SL_n(R, I)$ for any ideal $I$ satisfying $\sr(I)\leqslant n-1$.

Applying \cref{thm:DennisVaserstein} and proving a more accurate version of \cite[Proposition~1]{Nik07} in the special case $\Phi=\rD_\ell$, we obtain the following result.
\begin{thm}\label{thm:spin-sln-prod}
Assume that $\sr(I) \leqslant 2$. Then the group $\Epin_{2\ell}(R, I)=\E(\rD_\ell, R, I)$ is a product of at most $9$ conjugates of its (regularly embedded) subgroups of type $\rA_{\ell-1}$.
\end{thm}

The rest of the article is organized as follows.
In sections \ref{sec:prelim}--\ref{sec:stability-conditions} we introduce the principal notation and prove some basic facts pertaining to Chevalley groups over rings and the stability conditions.
In section \ref{sec:factorizations} we construct various parabolic factorizations and thus prove \cref{thm:srRI1,thm:DennisVaserstein,thm:SL2width}.
Finally, we treat the bounded generation and the product decompositions separately in section~\ref{sec:applications}, where \cref{thm:width,thm:spin-sln-prod} are proved.

\section{Preliminaries}\label{sec:prelim}
For any collection of subsets $H_1, \ldots, H_n$ of a group $G$ we denote by $H_1\ldots H_n$ their Minkowski set-product, 
i.\,e. the set consisting of arbitrary products $h_1\ldots h_n$ of elements $h_i\in H_i$. In particular, the equality
$G = H_1\cdot\ldots\cdot H_n$ means that every element $g\in G$ can be presented as a product $h_1\ldots h_n$, where $h_i\in H_i$ for each $i$.

\subsection{Root systems}\label{sec:rootsys}
Let $\Phi$ be a reduced irreducible root system of rank $\ell$ with a fixed basis of fundamental roots $\Pi=\{\alpha_1, \ldots, \alpha_l\}$.
We use the conventional numbering of basis vectors of $\Pi$ which follows Bourbaki (see~\cite[Table~1]{PSV98}).
Denote by $\Phi^+$ and $\Phi^-$ the subsets of positive and negative roots with respect to $\Pi$.
Let $\alpha_\mathrm{max}$ denote the maximal root of $\Phi$.
For a root $\alpha\in\Phi$ we denote by $m_i(\alpha)$ the $i$-th coefficient in the expansion of $\alpha$ with respect to $\Pi$, i.\,e. $\alpha = \sum_{i=1}^n m_i(\alpha) \alpha_i$.

@@

We denote by $(\alpha, \beta)$ the scalar product of roots and by $\langle \alpha, \beta\rangle$ the integer $2(\alpha, \beta)/(\beta, \beta)$.
Let $\varpi_1, \ldots, \varpi_\ell$ be the \emph{fundamental weights }of $\Phi$, i.\,e. vectors
defined by $\langle\varpi_i, \alpha_j^\vee\rangle=\delta_{i, j}, $ where $\alpha^\vee = \frac{\alpha}{2(\alpha, \alpha)}$.

Denote by $W(\Phi)$ the subgroup of isometries of $\Phi$ generated by all simple reflections $\sigma_\alpha$, $\alpha\in\Phi$. Recall that a simple reflection is defined by $\sigma_\alpha(\beta)=\beta-\langle\beta, \alpha \rangle\cdot \alpha$.
For a reductive subset $\Delta\subseteq \Phi$ denote by $W(\Delta)$ the subgroup of $W(\Phi)$ generated by $\sigma_\alpha$ for $\alpha\in\Delta$.

Let $J\subseteq \Pi$ be a subset of fundamental roots. 
For a root $\beta = \sum m_i(\beta)\alpha_i \in \Phi$ define \emph{$J$-shape} of $\beta$ (denoted $\shape(J, \beta)$) as $\sum_{i\in J} m_i(\beta) \alpha_i$.
Set
\begin{align*}
& \Delta_J = \{\alpha \in \Phi \mid \shape(J, \alpha)=0\}, \\
& \Sigma^\pm_J = \{\alpha \in \Phi \mid \shape(J, \alpha) \in \mathbb{Z}_{\geqslant0} \Phi^\pm \}, \\
& S_J^\pm = \Delta_J \sqcup \Sigma_J^\pm.
\end{align*}
Clearly, $\Delta_J$ is a reductive subset, while $S^\pm_J$ and $\Sigma^\pm_J$ are parabolic and special subsets, respectively.
For two disjoint subsets $I, J\subseteq \Pi$ one has 
\[ \Sigma^\pm_{I \cup J} = \Sigma^\pm_I\cup\Sigma^\pm_J, \quad \Delta_{I\cup J} = \Delta_I \cap \Delta_J. \]
We tend to omit curly braces in the above notations in the case when $J$ is a one or two element set, e.\,g., $\Delta_k=\Delta_{\{k\}}$ and $\Sigma_{i, j}^\pm=\Sigma_{\{i, j\}}^\pm$, etc.

@@

\subsection{Chevalley groups and their elementary subgroups} \label{sec:elementary}
Our exposition of Chevalley groups is standard and follows~\cite{Ta, S, St78, VP, Va86}.
We denote by $\G(\Phi, R)$ the simply-connected Chevalley group of type $\Phi$ over an arbitrary commutative ring $R$. For example, for each of the four classical series $\Phi=\rA_\ell, \rB_\ell, \rC_\ell, \rD_\ell$
the group $\G(\Phi, R)$ equals $\SL(\ell+1, R)$, $\Spin(2\ell+1, R)$, $\Sp(2\ell, R)$, $\Spin(2\ell, R)$, respectively.

Notice that from the uniqueness theorem of M.~Demazure (see~\cite[Corollaire~5.2]{SGA3}) 
it follows that the exceptional isomorphisms of small-rank groups (which are well-known in the theory of groups over fields) remain valid for Chevalley groups over arbitrary commutative rings.
In particular, there are isomorphisms of groups $\Spin(3, R)\cong\SL(2, R)$, $\Spin(5, R)\cong \Sp(4, R)$, $\Spin(6, R)\cong\SL(4, R)$, $\SL(2, R)\cong \Sp(2, R)$.

Recall from~\cite{St78, VP} that for $\alpha\in \Phi$, $\xi\in R$ one can define certain elements $x_{\alpha}(\xi)$ of $\G_{sc}(\Phi, R)$, called the {\it elementary root unipotents}.
These elements satisfy the well-known Steinberg relations:
\begin{align}
& \phantom{[}
x_\alpha(s) x_\alpha(t) = x_\alpha(s+t), \label{rel:add}\\
& [x_\alpha(s), x_\beta(t)] = \prod x_{i\alpha + j\beta}\left(N_{\alpha\beta ij}\, s^i t^j\right), \quad \alpha\neq-\beta, \quad N_{\alpha\beta ij}\in\mathbb{Z}. \label{rel:CCF}
\end{align} 
The indices $i$, $j$ in the above formula range over all positive integers such that $i\alpha + j\beta\in\Phi$.
The integers $N_{\alpha\beta ij}$ are called the \emph{structure constants} of the Chevalley group $\G(\Phi, R)$ and depend only on $\Phi$.
We refer the reader to~\cite[\S14]{VP} for basic properties of the structure constants.

For an additive subset $I\subseteq R$ we denote by $X_\alpha(I)$ the corresponding \emph{root subgroup} of level $I$, i.\,e. the subgroup consisting of all elementary root unipotents $x_\alpha(\xi)$, $\xi\in I$.

The subgroup $\E(\Phi, R)$, generated by all $X_\alpha(R)$, $\alpha\in\Phi$ is called the {\it elementary subgroup} of the Chevalley group $\G_{sc}(\Phi, R)$.

For $\varepsilon\in R^*$ set $w_\alpha(\varepsilon) = x_\alpha(\varepsilon) x_{-\alpha}(-\varepsilon^{-1}) x_{\alpha}(\varepsilon).$
If $\rk(\Phi)\geqslant 2$ the following relation is a consequence of \eqref{rel:add}--\eqref{rel:CCF}:
\begin{equation}\label{rel:R3}
w_\alpha(\varepsilon) x_{\beta}(\xi) w_\alpha(\varepsilon)^{-1} =
x_{\sigma\ssub{\alpha}\beta} \left(\eta_{\alpha, \beta}\cdot\varepsilon^{-\langle\beta, \alpha \rangle}\xi\right), \quad \varepsilon\in R^*, \ \xi\in R.
\end{equation}
where $\eta_{\alpha, \beta}=\pm 1$. The coefficients $\eta_{\alpha, \beta}$ can be expressed through the structure constants (see~\cite[\S13]{VP}).
For a reductive subset $\Delta \subseteq \Phi$ denote by $\widetilde{W}(\Delta)$ the \emph{extended Weyl group}, i.\,e. the subgroup of $\E(\Phi, R)$ generated by all $w_{\alpha}(1)$, $\alpha \in \Delta$.
\begin{lemma} \label{lemma:weylfacts} Let $\Phi$ be an irreducible root system and let $I$ be an ideal of $R$. 
\begin{lemlist}
\item \label{item-trans1} For every two roots $\alpha, \beta \in \Phi$ of the same length there exists $w \in \widetilde{W}(\Phi)$ such that $X_{\alpha}(I)^w = X_\beta(I)$.
\item \label{item-egen} The elementary group $\E(\Phi, R)$  is generated as an abstract group by the root subgroups $X_{\alpha}(R)$, $\alpha \in \Pi \cup -\Pi$.
\item \label{item-trans2} Assume that $\alpha_s\in \Pi$ is a fundamental root of $\Phi$ such that $m_s(\alpha_\mathrm{max}) = 1$. 
                          Then for $\alpha, \beta \in \Sigma^\pm_s$ of the same length there exists $w\in \widetilde{W}(\Delta_s)$ such that $X_\alpha(I)^w = X_\beta(I)$. \end{lemlist}
\end{lemma}
\begin{proof}
The first statement of the lemma follows from identity~\eqref{rel:R3} and the well-known fact that the Weyl group $W(\Phi)$ acts transitively on the set of roots of the same length.

>>>>

\documentclass[11pt]{amsart}
\usepackage{amscd, amsmath, amssymb, amsthm, amsfonts, amstext, verbatim, enumitem, graphicx, mathtools, xfrac, tikz-cd, microtype, nameref, thmtools}
\usepackage[breaklinks=true]{hyperref}
\usepackage[capitalize]{cleveref}

@@

%\oddsidemargin 5mm
%\marginparwidth 5mm
%\topmargin 0mm
%\textheight 225mm
%\textwidth 165mm
%\headheight 0mm
%\headsep 10mm
%\footskip 5mm

\usepackage[margin=2cm]{geometry}

@@

\title[Decompositions of congruence subgroups of Chevalley groups]{Decompositions of congruence subgroups \\ of Chevalley groups}
\keywords {parabolic decomposition, bounded generation, $\SL_2$-factorization, subsystem factorization, product decomposition. {\em Mathematical Subject Classification (2010):} 20G40, 20G35, 19B14}
\author{Sergey Sinchuk}
\author{Andrei Smolensky}
\email{sinchukss {\it at} yandex.ru; andrei.smolensky {\it at} gmail.com}
\thanks{The authors of the present paper acknowledge the financial support of the Russian Science Foundation grant 14-11-00297}

@@

\begin{abstract} 
We formulate and prove relative versions of several classical decompositions known in the theory of Chevalley groups over commutative rings.
We apply them to construct new factorizations in terms of $\SL_2$ and $\SL_n$ subgroups and to obtain upper estimates on the width of principal congruence subgroups in terms of Tits---Vaserstein generators.
Some of our results are new even in the absolute case and were previously studied only for groups over finite fields.
\end{abstract}

@@

\section{Introduction}\label{sec:intro}
The aim of the present paper is to extend various decomposition of Chevalley groups and their elementary subgroups to their principal congruence subgroups and relative elementary subgroups. We focus on the so-called ``parabolic decompositions'', i.e. decompositions involving parabolic and unipotent factors. The main result is the following

\begin{restatable*}[Dennis---Vaserstein decomposition]{thm}{ThmDennisVaserstein}\label{thm:DennisVaserstein}
Let $\Phi \neq \rE_8$ be a reduced irreducible root system of rank $\ell\geqslant 2$ and let $\{ r, s \}$ be a pair of distinct terminal vertices of the Dynkin diagram of $\Phi$. Denote the distance between $r$ and $s$ by $d$.
Assume that $\Phi$, $I$ and $\{r, s\}$ satisfy either of the following assumptions:
\begin{thmlist}
 \item $\sr(I) \leqslant d$ for classical $\Phi$,
 \item $\sr(I) \leqslant d$ for $\Phi=\rE_\ell$ with $\{r, s\} = \{ 2, \ell \}$,
 \item $\asr(I) \leqslant d$ for $\Phi=\rE_\ell$ with $\{r, s\} = \{1, \ell \}$,
 \item $\asr(I) \leqslant d$ for $\Phi=\rF_4$.
\end{thmlist}
Then $\E(\Phi, R, I) = \EP_r(R, I) \cdot \U(\Sigma^-_r \cap \Sigma^-_s, I) \cdot \EP_s(R, I).$
\end{restatable*}
Even in the special case $I=R$ the above theorem contains several new cases as compared to~\cite[Theorems~2.5 and~4.1]{St78}.
The term ``Dennis---Vaserstein decomposition'' is due to A.~Suslin and M.~Tulenbaev \cite[Lemma~2.1]{ST76}.
The condition $\asr(I) \leq d$ is a relative version of the \emph{absolute stable range} condition of D.~Estes and J.~Ohm, which depends only on the ideal $I$ (see section~\ref{sec:rel-asr}).

We also prove the relative analogues of Gauss and Bass---Kolster decompositions (\cref{thm:srRI1,thm:BassKolster}).

We apply these decompositions to several problems. First, we investigate the factorizations of Chevalley groups in terms of their subsystem subgroups. One of such factorizations if a decomposition into the product of subgroups isomorphic to $\SL(2, R, I)$. We show in \cref{thm:SL2width} that for classical groups over rings satisfying assumptions close to the stable rank $\leqslant2$ there exist such factorizations of length having the leading term $3|\Phi^+|$ in case $\Phi=\rA_\ell, \rC_\ell$ and $4|\Phi^+|$ otherwise.

Another is the decomposition in terms of subgroups of type $\rA_\ell$ of submaximal rank. We show that Dennis---Vaserstein decomposition can be of great help in this problem. Namely, we prove the following
\begin{restatable*}{thm}{ThmSpinSLnProd}\label{thm:spin-sln-prod}
Assume that $\sr(I) \leqslant 2$. Then the group $\Epin(2\ell, R, I)=\E(\rD_\ell, R, I)$ is a product of at most $9$ regularly embedded subgroups of type $\rA_{\ell-1}$.
\end{restatable*}

Then we obtain the upper bounds on the width of $\E(\Phi, R, I)$ in terms of Vaserstein---Tits generators for various rings (\cref{thm:width}).

Let us now fix the notations used in the paper.
Let $\Phi$ be a reduced irreducible root system of rank $\ell$ with a fixed basis of fundamental roots $\Pi=\{\alpha_1, \ldots, \alpha_l\}$.
We use the conventional numbering of basis vectors of $\Pi$ which follows Bourbaki (see~\cite[Table~1]{PSV98}).

For a root $\alpha\in\Phi$ we denote by $m_i(\alpha)$ the $i$-th coefficient in the expansion of $\alpha$ with respect to $\Pi$, i.\,e. $\alpha = \sum_{i=1}^n m_i(\alpha) \alpha_i$.

@@

We denote by $(\alpha, \beta)$ the scalar product of roots and by $\langle \alpha, \beta\rangle$ the integer $2(\alpha, \beta)/(\beta, \beta)$.
Let $\varpi_1, \ldots, \varpi_\ell$ be the \emph{fundamental weights }of $\Phi$.

Denote by $W(\Phi)$ the subgroup of isometries of $\Phi$ generated by all simple reflections $\sigma_\alpha$, $\alpha\in\Phi$.
For a reductive subset $\Delta\subseteq \Phi$ denote by $W(\Delta)$ the subgroup of $W(\Phi)$ generated by $\sigma_\alpha$ for $\alpha\in\Delta$.

Let $J\subseteq \Pi$ be a subset of fundamental roots. Define
\begin{align*}
& \shape(J, \beta)=\sum\nolimits_{i\in J} m_i(\beta) \alpha_i, \\
& \Delta_J = \{\alpha \in \Phi \mid \shape(J, \alpha)=0\}, \\
& \Sigma^\pm_J = \{\alpha \in \Phi \mid \shape(J, \alpha) \in \mathbb{Z}_{\geqslant0} \Phi^\pm \}, \\
& S_J^\pm = \Delta_J \sqcup \Sigma_J^\pm.
\end{align*}
Clearly, $\Delta_J$ is a reductive subset, while $S^\pm_J$ and $\Sigma^\pm_J$ are parabolic and special subsets, respectively.
For two disjoint subsets $I, J\subseteq \Pi$ one has 
\[ \Sigma^\pm_{I \cup J} = \Sigma^\pm_I\cup\Sigma^\pm_J, \quad \Delta_{I\cup J} = \Delta_I \cap \Delta_J. \]
We tend to omit curly braces in the above notations when $J$ is a one or two element set, e.\,g., $\Delta_k=\Delta_{\{k\}}$ and $\Sigma_{i, j}^\pm=\Sigma_{\{i, j\}}^\pm$, etc.

@@

Let $\G(\Phi, R)$ be the simply-connected Chevalley group of type $\Phi$ over an arbitrary commutative ring $R$ and let $\E(\Phi, R)$ be its elementary subgroup, generated by the elementary root unipotents $x_\alpha(\xi)$, $\alpha\in\Phi$, $\xi\in R$, see~\cite{VP, St78, S}.
For an ideal $I \trianglelefteq R$ we denote by $\G(\Phi, R, I)$ the \emph{relative Chevalley group} (alias the \emph{principal congruence subgroup}) of level $I$.

The \emph{relative elementary Chevalley subgroup} $\E(\Phi, R, I) \leqslant \G(\Phi, R, I)$ is defined as the normal closure in $\E(\Phi, R)$ of the subgroup $\E(\Phi, I)$ generated by the set $\mathcal{X} = \{x_\alpha(\xi) \mid \alpha\in \Phi, \ \xi \in I \}$.

Set $z_\alpha(s, \xi) = x_{\alpha}(s)^{x_{-\alpha}(\xi)}$, $s\in I$, $\xi\in R$.
It is a classical result of J.~Tits and L.~Vaserstein (see~\cite{Tits76}, \cite[Theorem~2]{Va86}) that the relative elementary subgroup $\E(\Phi, R, I)$ is generated by the set of elements $\mathcal{Z}(\Phi)$.
Here for a subset of roots $S\subseteq\Phi$ we denote by $\mathcal{Z}(S)$ the set
\[ \mathcal{Z}(S) =  \mathcal{X} \cup \{ z_\alpha(s, \xi) \mid s\in I, \ \xi \in R, \ \alpha \in S\}.\]
\begin{lemma}[{\cite[Corollary~3.3]{S}}]\label{lemma:Stepanov-ideal}
Let $\Phi$ be a root system of rank $\geqslant2$, let $R$ be a commutative ring and $I\trianglelefteq R$ be its ideal.
If $\Phi\neq\rC_\ell$ then $\E\left(\Phi, R, I^2\right)\leqslant\E(\Phi, I)$, otherwise $\E\left(\Phi, R, II^{\indexbox{2}}\right)\leqslant\E(\Phi, I)$.
\end{lemma}
Here $I^{\indexbox{2}}$ denotes the ideal generated by the squares $a^2$, where $a\in I$.

%The following recent result of A.~Stepanov served as a starting point for the present paper.
%The notation for the subgroups involved in the statements of the results below is introduced properly in section~\ref{sec:prelim}.
%\begin{prop}[{\cite[Theorem~3.4]{S}}] \label{prop:Stepanov}
%Let $\Phi$ be a reduced irreducible root system of rank $\geq 2$, let $S \subset \Phi$ be an arbitrary parabolic subset of roots, and $\Sigma_S$ be the special part of $S$. The relative elementary subgroup $\E(\Phi, R, I)$ is generated as an abstract group by the set $\mathcal{Z}(\Sigma_S)$. \end{prop}

%The purpose of this paper is two-fold. 
%First, we obtain relative analogues of several classical parabolic factorizations of Chevalley groups such as the Gauss, Bass---Kolster and Dennis---Vaserstein decompositions.
%Then we apply these factorizations to the study of various bounded generation properties of Chevalley groups over rings.
%
%Our first result is a relative version of the Gauss decomposition formulated in terms of the \emph{stable rank} of the ideal $I$ (see section~\ref{sec:stability-conditions} for the definition)
%\begin{thm}\label{thm:srRI1}
%Let $\Phi$ be a root system, let $I$ be an ideal of an arbitrary commutative ring $R$, and assume that $\sr(I)=1$.
%Then the relative elementary Chevalley group $\E(\Phi, R, I)$ admits the following decomposition:
%\[ \E(\Phi, R, I) = \Hh(\Phi, R, I) \cdot \U(\Phi, I) \cdot \U(\Phi^-, I) \cdot \U(\Phi, I). \]
%\end{thm}
%In the special case when $I$ is contained in the Jacobson radical of $R$, \cref{thm:srRI1} boils down to~\cite[Proposition~2.3]{Abe76}. 
%We also note that the absolute case (i.\,e. $I=R$) of the theorem was treated by the second named author in~\cite{Sm12}.
%
%The next main result is the relative Dennis---Vaserstein decomposition, which holds for a broader class of rings as compared to \cref{thm:srRI1}.
%\begin{thm}\label{thm:DennisVaserstein}
%Let $\Phi \neq \rE_8$ be a reduced irreducible root system of rank $\ell\geqslant 2$ and let $\{ r, s \}$ be a pair of distinct terminal vertices of the Dynkin diagram of $\Phi$. Denote the distance between $r$ and $s$ by $d$.
%Assume that $\Phi$, $I$ and $\{r, s\}$ satisfy either of the following assumptions:
%\begin{thmlist}
% \item $\sr(I) \leqslant d$ for classical $\Phi$,
% \item $\sr(I) \leqslant d$ for $\Phi=\rE_\ell$ with $\{r, s\} = \{ 2, \ell \}$,
% \item $\asr(I) \leqslant d$ for $\Phi=\rE_\ell$ with $\{r, s\} = \{1, \ell \}$,
% \item $\asr(I) \leqslant d$ for $\Phi=\rF_4$.
%\end{thmlist}
%Then $\E(\Phi, R, I) = \EP_r(R, I) \cdot \U(\Sigma^-_r \cap \Sigma^-_s, I) \cdot \EP_s(R, I).$ \end{thm}
%Even in the special case $I=R$ the above theorem contains several new cases as compared to~\cite[Theorems~2.5 and~4.1]{St78}.
%The term ``Dennis---Vaserstein decomposition'' is due to A.~Suslin and M.~Tulenbaev (cf.~\cite[Lemma~2.1]{ST76}).
%
%The condition $\asr(I) \leq d$ used in the statement of \cref{thm:DennisVaserstein} is a relative version of the \emph{absolute stable range} condition of D.~Estes and J.~Ohm, which depends only on the ideal $I$ (see section~\ref{sec:rel-asr}).
%
%Let $G$ be an abstract group with a generating set $X \subset G$. 
%We denote by $W(G, X)$ the \emph{width of $G$ with respect to $X$}, i.\,e. the smallest natural number $N$ such that every element of $G$ is a product of at most $N$ elements of $X$.
%
%Our first application of the parabolic factorizations is the following theorem, which gives an upper bound on the width of $\E(\Phi, R)$ in terms of Tits generators. 
%Thus, it is a quantitative analogue of \cref{prop:Stepanov}
%and a relative analogue of the results of~\cite{Tavgen91, VseUnitrZ1p, VavSmSuUnitrEng}.
%\begin{thm}\label{thm:width} Let $I$ be an ideal of a commutative ring $R$.
%\begin{thmlist}
%\item If $R=\mathcal{O}_S$ is a Dedekind ring of arithmetic type posessing a real embedding and $\Phi$ is classical of rank $\geqslant2$, then 
%$W(\G(\Phi, R, I), \mathcal{Z}(\Sigma_\ell^-))$ is finite;
%\item If $\sr(I) = 1$ and $\Phi$ is an arbitrary irreducible root system, then 
%\[W(\E(\Phi, R, I), \mathcal{Z}(\Pi))\leqslant 3|\Phi^+|+2\rk(\Phi)-1;\]
